### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding figures for nonlinear annalysis

\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.2}{
      \input{chapters/nonlinear_problem/backward_facing_step.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the backward facing step
      benchmark.} \textit{Zero velocity boundary conditions are
      applied at the top and bottom of the domain while the outflow
      boundary condition on the right boundary is represented by zero
      stress tensor components in the direction of the flow. For the
      inlet conditions, the left boundary is split such that the top
      half has a fully formed parabolic flow profile and the bottom
      half has a zero velocity condition, simulating flow over a
      step. The $H$ parameter given in the parabolic inlet condition
      is the width of the inlet.}}
  \label{fig:backward_facing_step}
\end{figure}

@@

Given the difficulty of these benchmark problems, the linear solve at
each Newton step was preconditioned with an algebraic multigrid
method\footnote{The ML algebraic multigrid library in the Trilinos
  scientific computing libraries \citep{heroux_overview_2005} was used
  for all problems.}, a common technique for preconditioning these
equations \citep{ghia_high-re_1982,evans_enhanced_2007}. Without
multigrid preconditioning, the residual in MCSA was observed to
stagnate. Typically this happens because the Richardson iteration and
Neumann series on which the Monte Carlo method is based will quickly
damp the higher order error modes on the scale of the computational
grid used to solve the problem but will reduce lower order error modes
at a considerably lower rate. In terms of Monte Carlo, this stagnation
manifests itself in histories that simply do no terminate and
therefore no solution is acquired.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Newton-Krylov
method. For each benchmark, we will analyze the iterative performance
of both the Newton solver and the linear solver used to compute the
correction vector at each step. In addition, timing results will be
discussed.

@@

At each Rayleigh number used for the thermal convection cavity
problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
the number of Newton iterations required for convergence for each
method. Limiting the number of nonlinear iterations required for
solution is critical for efficiency as nonlinear function evaluations,
Jacobian generation, and preconditioner construction are typically
required at every Newton iteration. For all Rayleigh numbers except
\sn{1}{5} the number of Newton iterations required to converge is the
same while the \sn{1}{5} result required an extra FANM iteration. This
is due to the fact that the convergence of the linear residual has a
strong effect on the convergence of the nonlinear residual. For
nonlinear calculations, this is much stronger than the effect on the
convergence of the eigenvalue in the $SP_N$ computations where we
always observed that identical numbers of k-eigenvalue iterations were
required to converge each variation of the probem.

@@

Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
behavior for the \sn{1}{5} Rayleigh number case that required the
extra FANM iteration. We see that the extra iteration comes from
slightly slower quadratic convergence with FANM for this particular
case. The difference in convergence of the nonlinear problem comes
from the fact that the same multigrid preconditioner parameters are
used for this case as for the \sn{1}{3} and \sn{1}{4} Rayleigh number
cases for the linear solutions. Using the same preconditioner for the
harder \sn{1}{5} case means that the problem is still growing in
difficulty even though it is preconditioned in an identical
manner. The more ill-conditioned the linear model is, the closer the
linear residual is converged to the requested tolerance whereas a
better conditioned problem may produce a linear solution that
overshoots the specified tolerance provided by the forcing term. When
oversolving the linear model is not an issue, this results in faster
convergence of the nonlinear problem.

For the \sn{1}{5} Rayleigh number case, this means that the GMRES
solver at each Newton step is converging the linear residual to a
smaller tolerance than the MCSA solver, even though they have
approximately the same forcing terms to begin with. As the Rayleigh
number increases, the problem becomes more difficult to solve and
therefore a more agressive preconditioner should maintain iterative
performance of the nonlinear method. This was observed to be true for
the \sn{1}{6} Rayleigh number case. For this calculation, the number
of multigrid cycle applications was bumped up to 3 from 1 and FANM
converged in the same number of nonlinear iterations as the
Newton-Krylov method.

@@

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/ra_1e5_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the \sn{1}{5} Rayleigh number case.} \textit{The
      FANM calculation required an additional nonlinear iteration to
      converge.}}
  \label{fig:ra1e5_convergence}
\end{figure}

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solvers preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme. Even the
\sn{1}{5} Rayleigh number case that required more Newton iterations
required 5 fewer MCSA iterations to converge than GMRES
iterations. If the performance of the MCSA implementation used here
were to be potentially optimized in the future, this improved
iterative performance could also mean improved time to solution.

@@

@@

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/re700_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the 700 Reynolds number case.} \textit{The FANM
      calculation converged in 4 fewer iterations than the
      Newton-Krylov calculation.}}
  \label{fig:re700_convergence}
\end{figure}

@@

These iterative results again point out the fundamental flaw of
requiring a spectral radius of less than unity for the MCSA iteration
to converge at each Newton step. As the problem becomes more dominated
by inertial forces, the resulting linear models become increasingly
stiff and difficult to precondition. As a result, convergence requires
many more iterations than needed for the flow problem driven by
convection in the previous benchmark if convergence is achieved at
all. Much of this can be attributed to the physics at play in each of
these benchmarks. For the thermal convection cavity benchmark, the
forces at work in that problem were more effectively resolved by the
Monte Carlo method as they are more diffusive in character than the
flow driven by intertial forces in the lid driven cavity problem. This
physics-based observation was also true for the $SP_N$ equations where
the diffusion dominated physics in that case gave better convergence
results for MCSA than GMRES.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{Newton-Krylov Iterations}&
      \multicolumn{1}{c}{FANM Iterations}\\
      \hline
      200 & 10 & 9 \\
      400 & 15 & 14 \\
      600 & 10 & 10 \\
      800 & 19 & 20 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:step_nonlinear_iter_comparison}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}\\
      \hline
      200 & 24 & 13 \\
      400 & 23 & 17 \\
      600 & 18 & 12 \\
      800 & 30 & 52 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step total sum of linear
      iterations for all nonlinear iterations.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:step_linear_iter_comparison}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
      \hline
      200 & 400 \\
      400 & 593 \\
      600 & 825 \\
      800 & 1057 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step Newton-Krylov speedup
      over FANM method.} \textit{Speedup values are rounded to the
      nearest integer. The results presented here were obtained from
      the benchmark verification calculations.}}
  \label{tab:step_speedup_comparison}
\end{table}

@@

As a final overall comparison, we tabulate the results from the
benchmark calculations over the three performance metrics: nonlinear
iterations, linear iterations, and CPU
time. Tables~\ref{tab:benchmark_nonlinear_comparison},
\ref{tab:benchmark_linear_comparison}, and
\ref{tab:benchmark_time_comparison} give the results. In these tables,
the calculations that procduced the best performing results are
tallied. In the case of a tie, both are tallied. Looking at the
tabulated results, we see marginally better performance for FANM in
terms of nonlinear iterations with 1 extra case outperforming the
Newton-Krylov method although this is not conclusive enough to claim
superior performance converging the nonlinear residual with FANM. As
observed for each benchmark, the timing results were always in favor
of the Newton-Krylov method due to the effects of the explict
preconditioning strategy.

>>>>

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.2}{
      \input{chapters/nonlinear_problem/backward_facing_step.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the backward facing step
      benchmark.} \textit{Zero velocity boundary conditions are
      applied at the top and bottom of the domain while the outflow
      boundary condition on the right boundary is represented by zero
      stress tensor components in the direction of the flow. For the
      inlet conditions, the left boundary is split such that the top
      half has a fully formed parabolic flow profile and the bottom
      half has a zero velocity condition, simulating flow over a
      step. The $H$ parameter given in the parabolic inlet condition
      is the width of the inlet.}}
  \label{fig:backward_facing_step}
\end{figure}

@@

Given the difficulty of these benchmark problems, the linear solve at
each Newton step was preconditioned with an algebraic multigrid
method\footnote{The ML algebraic multigrid library in the Trilinos
  scientific computing libraries \citep{heroux_overview_2005} was used
  for all problems.}, a common technique for preconditioning these
equations \citep{ghia_high-re_1982,evans_enhanced_2007} with the
preconditioner reconstructed at every Newton step for the new
Jacobian. Without multigrid preconditioning, the residual in MCSA was
observed to stagnate. Typically this happens because the Richardson
iteration and Neumann series on which the Monte Carlo method is based
will quickly damp the higher order error modes on the scale of the
computational grid used to solve the problem but will reduce lower
order error modes at a considerably lower rate. In terms of Monte
Carlo, this stagnation manifests itself in histories that simply do no
terminate and therefore no solution is acquired.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Newton-Krylov
method. For each benchmark, we will analyze the iterative performance
of both the nonlinear Newton solver and the linear solver used to
compute the Newton correction vector at each step. Limiting the number
of nonlinear iterations required for solution is critical for
efficiency as nonlinear function evaluations, Jacobian generation, and
preconditioner construction are typically required at every Newton
iteration. In addition, timing results will also be discussed.

@@

At each Rayleigh number used for the thermal convection cavity
problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
the number of Newton iterations required for convergence for each
method.  For all Rayleigh numbers except \sn{1}{5} the number of
Newton iterations required to converge is the same while the \sn{1}{5}
result required an extra FANM iteration. This is due to the fact that
the convergence of the linear residual has a strong effect on the
convergence of the nonlinear residual. For nonlinear calculations,
this is much stronger than the effect on the convergence of the
eigenvalue in the $SP_N$ computations where we always observed that
identical numbers of k-eigenvalue iterations were required to converge
each variation of the probem.

@@

Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
behavior for the \sn{1}{5} Rayleigh number case that required the
extra FANM iteration. We see that the extra iteration comes from
slightly slower quadratic convergence with FANM for this particular
case. The difference in convergence of the nonlinear problem comes
from the fact that the same multigrid preconditioner parameters are
used for this case as for the \sn{1}{3} and \sn{1}{4} Rayleigh number
cases for the linear solutions. Using the same preconditioner for the
harder \sn{1}{5} case means that the problem is still growing in
difficulty even though it is preconditioned in an identical
manner.

For the \sn{1}{5} Rayleigh number case, this means that the GMRES
solver at each Newton step is converging the linear residual to a
smaller tolerance than the MCSA solver as the forcing terms adapt to
the convergence of the method over the course of nonlinear
iterations. As the Rayleigh number increases, the problem becomes more
difficult to solve and therefore a more agressive preconditioner
should maintain iterative performance of the nonlinear method. This
was observed to be true for the \sn{1}{6} Rayleigh number case. For
this calculation, the number of multigrid cycle applications was
bumped up to 3 from 1 and FANM converged in the same number of
nonlinear iterations as the Newton-Krylov method.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/ra_1e5_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the \sn{1}{5} Rayleigh number case.} \textit{The
      FANM calculation required an additional nonlinear iteration to
      converge.}}
  \label{fig:ra1e5_convergence}
\end{figure}

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solvers preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case where a constant forcing term of \sn{1}{-4} was kept, this
difference is most drastic with FANM requiring only 64\% of the linear
iterations required by the Newton-Krylov scheme with the linear solver
per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. If the performance of
the MCSA implementation used here were to be potentially optimized in
the future, this improved iterative performance could also mean
improved time to solution.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/convection_ra1e6_iters.pdf}
  \end{center}
  \caption{\textbf{Linear solver iterations as a function of Newton
      iteration for the \sn{1}{6} Rayleigh number case.} \textit{MCSA
      shows better iterative performance than GMRES at all but the
      second nonlinear iteration. The forcing term was fixed at
      \sn{1}{-4} at all nonlinear iterations.}}
  \label{fig:ra1e6_linear_iters}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/re700_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the 700 Reynolds number case.} \textit{The FANM
      calculation converged in 4 fewer iterations than the
      Newton-Krylov calculation.}}
  \label{fig:re700_convergence}
\end{figure}

@@

As an additional means of analysis for the linear solver iteration
results, we will look at a case where MCSA performed poorly compared
to GMRES. Consider the linear iterations as a function of nonlinear
iteration given in Figure~\ref{fig:driven_re500_iters} and the forcing
term as a function of nonlinear iteration given in
Figure~\ref{fig:driven_re500_forcing} for the 500 Reynolds number
case. At the later Newton iterations where the region of fast
quadratic convergence is reached, we see that MCSA required many more
iterations to converge than GMRES, accounting for the majority of the
extra iterations. However, we see in the forcing term data that
forcing terms at later Newton iterations for MCSA were much smaller
than those for GMRES. This means that MCSA was converged to a tighter
tolerance at these iterations and therfore required the extra
iterations.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_iters.pdf}
  \end{center}
  \caption{\textbf{Linear solver iterations as a function of Newton
      iteration for the 500 Reynolds number case.} \textit{More MCSA
      iterations were required to converge the linear models at later
      Newton iterations due to smaller forcing terms.}}
  \label{fig:driven_re500_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_forcing.pdf}
  \end{center}
  \caption{\textbf{Forcing terms as a function of Newton
      iteration for the 500 Reynolds number case.} \textit{Once the
      region of fast quadratic convergence is reached, MCSA forcing
      terms were much larger.}}
  \label{fig:driven_re500_forcing}
\end{figure}

@@

The backward step problem exhibited different performance
characteristics than the other two benchmark problems, likely due to
the different geometry and boundary
conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
nonlinear iteration results for this benchmark. At low Reynolds
numbers, FANM converges in fewer nonlinear iterations than the
Newton-Krylov solver. However, as the problem becomes more difficult
to solver at higher rates of flow, the Newton-Krylov solver eventually
converges in fewer nonlinear iterations. We see the same behavior as
well in the linear solver iteration data given by
Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
numbers, FANM required significantly fewer linear solver iterations to
converge the problem while the ill-conditioned case at a Reynolds
number of 500 is observed to use many more MCSA iterations.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{Newton-Krylov Iterations}&
      \multicolumn{1}{c}{FANM Iterations}\\
      \hline
      200 & 10 & 9 \\
      300 & 15 & 14 \\
      400 & 10 & 10 \\
      500 & 19 & 20 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:step_nonlinear_iter_comparison}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}\\
      \hline
      200 & 24 & 13 \\
      300 & 23 & 17 \\
      400 & 18 & 12 \\
      500 & 30 & 52 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step total sum of linear
      iterations for all nonlinear iterations.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:step_linear_iter_comparison}
\end{table}

Unlike the lid driven cavity benchmark, the extra linear solver
iterations are not due to the smaller forcing terms but are rather
truly a function of the ill-conditioning of the Jacobian matrices
produced at each Newton step.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
      \hline
      200 & 400 \\
      300 & 593 \\
      400 & 825 \\
      500 & 1057 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step Newton-Krylov speedup
      over FANM method.} \textit{Speedup values are rounded to the
      nearest integer. The results presented here were obtained from
      the benchmark verification calculations.}}
  \label{tab:step_speedup_comparison}
\end{table}

@@

As a final overall comparison, we tabulate the results from the
benchmark calculations over the three performance metrics: nonlinear
iterations, linear iterations, and CPU
time. Tables~\ref{tab:benchmark_nonlinear_comparison},
\ref{tab:benchmark_linear_comparison}, and
\ref{tab:benchmark_time_comparison} give the results. In these tables,
the calculations that procduced the best performing results are
tallied. In the case of a tie, both are tallied. Looking at the
tabulated results, we see marginally better performance for FANM in
terms of nonlinear iterations with 1 extra case outperforming the
Newton-Krylov method. As observed for each benchmark, the timing
results were always in favor of the Newton-Krylov method due to the
effects of the explict preconditioning strategy.
